Algebraic Graphs with Class

### 2021-05-08 paper/paper.tex
Finish first draft of paper

\maketitle
\begin{abstract}
  \cite{mokhov2017algebraic} 
\end{abstract}

@@

\subsection{Algebraic Structure}\label{sec:algebra}
Before turning to more advanced graph constructions we will observe the
algebraic structure of the primitives. It turns out that overlay and connect
almost form a semiring over graphs:
\begin{itemize}
\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent commutative monoid
\item $(\mathcal{G}, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$
\end{itemize}
The shared identity and the missing annihilating zero for $\connect$
(i.e.\ $x \connect 0 = 0$) are the only difference. There is also the following
\textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
\[
  x \connect y \connect z = x \connect y \overlay x \connect z \overlay y \connect z
\]
Using decomposition we can actually prove some of the above properties, leading
to the following minimal set of axioms characterizing directed graphs:
\begin{itemize}
\item $(G, \overlay)$ is a commutative semigroup
\item $(G, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$, i.e.\ $x \connect (y \overlay z)
  = x \connect y \overlay x \connect z$ and $(x \overlay y) \connect z = x
  \connect z \overlay y \connect z$
\item decomposition: $x \connect y \connect z = x \connect y \overlay x
  \connect z \overlay y \connect z$ 
\end{itemize}
\begin{figure*}
  \begin{subfigure}[b]{0.4\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-distributivity.pdf}}
    \caption{Distributivity: $1 \connect (2 \overlay 3) = 1 \connect 2 \overlay
      1 \connect 3$ }
  \end{subfigure}
  \hspace{12mm}
  \begin{subfigure}[b]{0.5\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
    \caption{Decomposition: $1 \connect 2 \connect 3 = {1 \connect 2} \overlay
      1 \connect 3 \overlay 2 \connect 3$}
  \end{subfigure}
  \vspace{-1mm}
  \caption{Two axioms of the algebra of graphs. Illustrations from~\cite{mokhov2017algebraic}.\label{fig:axioms}}
\end{figure*}
One can easily check that the definitions of the primitives given
in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
completeness algebraic graphs. For any graph $G=(V,E)$ the following function
builds it using the four construction primitives:
\begin{minted}{haskell}
  graph :: Graph g => [Vertex g] -> [(Vertex g, Vertex g)] -> g
  graph vs es = overlay (vertices vs) (edges es)
\end{minted}
The \hs{edges} function generalizes \hs{edge} to a list of edges:
\begin{minted}{haskell}
  edges :: Graph g => [(Vertex g, Vertex g)] -> g
  edges = foldr overlay empty . map (uncurry edge)
\end{minted}
This was also formally proven in Isabelle/HOL (see~\autoref{sec:verification}).

Mokhov also follows a standard approach of defining a partial order $(\preceq)$ for the
idempotent overlay operation as $x \preceq y \iff x \overlay y = y$. This is indeed
a partial order under the graph axioms (i.e.\ it is reflexive, antisymmetric and
transitive). In fact this partial order can actually be used as a definition for
the subgraph relation:
\[
  x \subseteq y \coloneqq x \preceq y \iff x \overlay y = y
\]

So far we only considered directed graphs. Undirected graphs can be
obtained from the same construction primitives as directed graphs. It suffices
to modify the underlying axioms. In the case of undirected graphs this can be
achieved by making connect commutative (then denoted by $\leftrightarrow$),
i.e.\ $x \leftrightarrow y = y \leftrightarrow x$. By introducing this axiom we
can prove some of the axioms of directed graphs. So undirected graphs are
characterized by the following minimal set of axioms:
\begin{itemize}
\item $(\mathcal{G}, \oplus)$ is a commutative semigroup
\item $\leftrightarrow$ is commutative and has $\eps$ as the identity
\item left distributivity: $x \leftrightarrow (y \oplus z) = x \leftrightarrow
  y \oplus x \leftrightarrow z$
\item left decomposition: $(x \leftrightarrow y) \leftrightarrow z = x
  \leftrightarrow y \oplus x \leftrightarrow z \oplus y \leftrightarrow z$
\end{itemize}
By choosing different additional axioms, other classes of graphs can be
characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
self-loop), transitive graphs (as in dependency graphs), and the combination
thereof; preorders. Combining undirected graphs and reflexive graphs yields
equivalence relations. It is even possible to define hypergraphs of arbitrary
(but fixed) order by replacing the decomposition law with another suitable
axiom~\cite{mokhov2017algebraic}.

@@

Equipped with these two instances let us now actually consider graph
constructions and transformations made possible with the core type class. Simple
examples include constructing a \textit{path} from a list of vertices, which
then can be easily extended to a \textit{circuit}:
\begin{minted}{haskell}
path :: Graph g => [Vertex g] -> g
path []  = empty
path [x] = vertex x
path xs = edges $ zip xs (tail xs)
%$
circuit :: Graph g => [Vertex g] -> g
circuit [] = empty
circuit xs = path (xs ++ [head xs])
\end{minted}
Mokhov also gives functions for constructing \textit{star} graphs (i.e.\ a
center vertex connected to a set of leaves) and for building graphs from
\hs{Tree}s and \hs{Forest}s from the \texttt{containers} library.

@@

Mokhov also goes on to introduce a \hs{Monad}-like \hs{newtype} wrapper. The
\hs{bind} in the context of graphs allows to replace each vertex with a
(possibly) empty subgraph. This can be used to both remove and split vertices.
\begin{minted}{haskell}
newtype GraphMonad a = M { bind :: forall g. Graph g => (a -> g) -> g }

@@

As the statements in the original paper were of very limited scope, we 
define and reason about more \textit{advanced} graph theoretic concepts: We
introduce (vertex) walks and the notion of reachability on algebraic graphs.
First impressions suggest that it is feasible to base more sophisticated
formalizations of graph theory on algebraic graphs. Although we suspect that
some additional axiom(s) may be required to make for \textit{interesting} graph
theory; as it stands the statement $\eps \neq \hs{vertex u}$ for a $u \in
\mathbb{V}$. This is further illustrated by the fact that we can instantiate the
algebraic graph locale with the unit type.

This is an interesting direction for future work for a related reason that the
type class is interesting in Haskell: The type class is the basis for 
polymorphic graph transformation library, whereas the locale can be
seen as a \textit{polymorphic} graph formalization library, as any lemma we can prove
using the locale axioms become available to a graph representation
instantiating it. This obviously becomes more interesting the more material is
available in the locale. Haslbeck et al.\ chose a similar approach to implement
and verify Kruskal's algorithm independent of the concrete graph
representation~\cite{Kruskal-AFP}.

\section{Deep Embedding}\label{sec:deep}
Up until this point we were mainly concerned with exploiting the algebraic
structure of graphs for polymorphic graph constructions and transformations, and
for formal verification. It turns out, however, that the deep embedding of the
construction primitives in a datatype is interesting in itself. Recall the
definition from the introduction:
\begin{minted}{haskell}
data Graph a = Empty
             | Vertex a
             | Overlay (Graph a) (Graph a)
             | Connect (Graph a) (Graph a)
\end{minted}
An instance for the type class can be defined easily, however, we cannot use the
derived \hs{Eq} instance for the datatype, as it would violate the graph axioms.
One solution is to fold the datatype into a concrete graph representation to
obtain a law-abiding \hs{Eq} instance:
\begin{minted}{haskell}
fold :: Graph g => Graph (Vertex g) -> g
fold Empty         = empty
fold (Vertex  x  ) = vertex x
fold (Overlay x y) = overlay (fold x) (fold y)
fold (Connect x y) = connect (fold x) (fold y)

instance Ord a => Eq (Graph a) where
  x == y = fold x == (fold y :: Relation a)
\end{minted}
This approach is also currently implemented in the library available on
\texttt{hackage}. However, this is not ideal for multiple reasons. Coupling the
datatype to some other graph representation merely for equality checks is
inconvenient at best. More interestingly algebraic graphs can compactly
represent dense graphs. Consider the graph \hs{clique [1..n]}, which is linear
in size in the deep embedding, while it is quadratic when stored as a
\hs{Relation}. Requiring to fully expand graphs to check equality voids this
advantage.
\begin{itemize}
\item compact representation
\item compare to different concrete/executable representations wrt.\ efficiency
  in time and space 
  \begin{itemize}
  \item sparse graphs (edge list)
  \item adjacency map
  \end{itemize}
\item performance (\href{https://github.com/haskell-perf/graphs}{haskell-perf})
\item reasoning in Isabelle/HOL \textrightarrow{} induction is great
\item algorithms directly on deep embedding? \textrightarrow{} future work/open
  question (Haskell implementation converts to adjacency map for algorithms)
\end{itemize}
\subsection{Quotient Type}\label{sec:quotient}
\begin{itemize}
\item deep embedding which satisfies axioms
\item additional lemmas
\item minimization~\cite{mcconnell2005linear}
  \begin{itemize}
  \item interesting for quotient type, disconnect from Noschinski/any other
    representation for equality, same applies in Haskell for Eq instance)
    \item interesting for algorithms (assuming we can exploit compact
      representation)
    \end{itemize}
\end{itemize}

\section{Conclusion}\label{sec:conclusion}
\begin{itemize}
\item summarize findings
\item restate open questions, future work
\item labeled graphs (outlook)
\item related work: other algebraic approaches (semiring on
  matrices~\cite{dolan2013fun}, relational
  algebra~\cite{berghammer2020relational})
\item usage examples~\cite{mokhov2019language,beaumont2017high}
\end{itemize}
\printbibliography
\end{document}

>>>>

\maketitle
\begin{abstract}
  We give an overview of algebraic graphs as introduced by
  Mokhov~\cite{mokhov2017algebraic}. Algebraic graphs have a small and safe core
  of construction primitives, alongside a set of axioms, characterizing an
  algebra of graphs. They can be used to elegantly implement a graph
  transformation library employing functional programming. We also review the
  suitability of algebraic graphs for formal verification using Isabelle/HOL.
\end{abstract}

@@

\subsection{Algebraic Structure}\label{sec:algebra}
Before turning to more advanced graph constructions we will observe the
algebraic structure of the primitives. It turns out that overlay and connect
almost form a semiring over graphs:
\begin{itemize}
\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent commutative monoid
\item $(\mathcal{G}, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$
\end{itemize}
The shared identity and the missing annihilating zero for $\connect$
(i.e.\ $x \connect 0 = 0$) are the only difference. There is also the following
\textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
\[
  x \connect y \connect z = x \connect y \overlay x \connect z \overlay y \connect z
\]
Using decomposition we can actually prove some of the above properties, leading
to the following minimal set of axioms characterizing directed graphs:
\begin{itemize}
\item $(G, \overlay)$ is a commutative semigroup
\item $(G, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$, i.e.\ $x \connect (y \overlay z)
  = x \connect y \overlay x \connect z$ and $(x \overlay y) \connect z = x
  \connect z \overlay y \connect z$
\item decomposition: $x \connect y \connect z = x \connect y \overlay x
  \connect z \overlay y \connect z$ 
\end{itemize}
\begin{figure*}
  \begin{subfigure}[b]{0.4\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-distributivity.pdf}}
    \caption{Distributivity: $1 \connect (2 \overlay 3) = 1 \connect 2 \overlay
      1 \connect 3$ }
  \end{subfigure}
  \hspace{12mm}
  \begin{subfigure}[b]{0.5\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
    \caption{Decomposition: $1 \connect 2 \connect 3 = {1 \connect 2} \overlay
      1 \connect 3 \overlay 2 \connect 3$}
  \end{subfigure}
  \vspace{-1mm}
  \caption{Two axioms of the algebra of graphs. Illustrations from~\cite{mokhov2017algebraic}.\label{fig:axioms}}
\end{figure*}
One can easily check that the definitions of the primitives given
in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
completeness of algebraic graphs. For any graph $G=(V,E)$ the following function
builds it using the four construction primitives:
\begin{minted}{haskell}
  graph :: Graph g => [Vertex g] -> [(Vertex g, Vertex g)] -> g
  graph vs es = overlay (vertices vs) (edges es)
\end{minted}
The \hs{edges} function generalizes \hs{edge} to a list of edges:
\begin{minted}{haskell}
  edges :: Graph g => [(Vertex g, Vertex g)] -> g
  edges = foldr overlay empty . map (uncurry edge)
\end{minted}
This was also formally proven in Isabelle/HOL (see~\autoref{sec:verification}).

Mokhov also follows a standard approach of defining a partial order $(\preceq)$ for the
idempotent overlay operation as $x \preceq y \iff x \overlay y = y$. This is indeed
a partial order under the graph axioms (i.e.\ it is reflexive, antisymmetric and
transitive). In fact this partial order can actually be used as a definition for
the subgraph relation, denoted by $\subseteq$:
\[
  x \subseteq y \coloneqq x \preceq y \iff x \overlay y = y
\]

So far we only considered directed graphs. Undirected graphs can be
obtained from the same construction primitives as directed graphs. It suffices
to modify the underlying axioms. In the case of undirected graphs this can be
achieved by making connect commutative (then denoted by $\leftrightarrow$),
i.e.\ $x \leftrightarrow y = y \leftrightarrow x$. By introducing this axiom we
can prove some of the axioms of directed graphs. Hence, undirected graphs are
characterized by the following minimal set of axioms:
\begin{itemize}
\item $(\mathcal{G}, \oplus)$ is a commutative semigroup
\item $\leftrightarrow$ is commutative and has $\eps$ as the identity
\item left distributivity: $x \leftrightarrow (y \oplus z) = x \leftrightarrow
  y \oplus x \leftrightarrow z$
\item left decomposition: $(x \leftrightarrow y) \leftrightarrow z = x
  \leftrightarrow y \oplus x \leftrightarrow z \oplus y \leftrightarrow z$
\end{itemize}
By choosing different additional axioms, other classes of graphs can be
characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
self-loop), transitive graphs (as in dependency graphs), and the combination
thereof; preorders. Combining undirected graphs and reflexive graphs yields
equivalence relations. It is even possible to define hypergraphs of arbitrary
(but fixed) order by replacing the decomposition law with another suitable
axiom~\cite{mokhov2017algebraic}.

@@

Equipped with these two instances let us now actually consider graph
constructions and transformations made possible with the core type class. Simple
examples include constructing a \textit{path} from a list of vertices, which
then can be easily extended to a \textit{circuit}:
\begin{minted}{haskell}
  path :: Graph g => [Vertex g] -> g
  path []  = empty
  path [x] = vertex x
  path xs = edges $ zip xs (tail xs)

  circuit :: Graph g => [Vertex g] -> g
  circuit [] = empty
  circuit xs = path (xs ++ [head xs])
\end{minted}
Mokhov also gives functions for constructing \textit{star} graphs (i.e.\ a
center vertex connected to a set of leaves) and for building graphs from
\hs{Tree}s and \hs{Forest}s from the \texttt{containers} library.

@@

Mokhov also goes on to introduce a \hs{Monad}-like \hs{newtype} wrapper. The
\hs{bind} in the context of graphs allows to replace each vertex with a
(possibly empty) subgraph. This can be used to both remove and split vertices.
\begin{minted}{haskell}
  newtype GraphMonad a = M { bind :: forall g. Graph g => (a -> g) -> g }

@@

As the statements in the original paper were of very limited scope, we 
define and reason about more \textit{advanced} graph theoretic concepts: We
introduce (vertex) walks and the notion of reachability on algebraic graphs.
First impressions suggest that it is feasible to base more sophisticated
formalizations of graph theory on algebraic graphs. Although we suspect that
some additional axiom(s) may be required to make for \textit{interesting} graph
theory; as it stands the statement $\eps \neq \hs{vertex u}$ for a $u \in
\mathbb{V}$ is unprovable. This is further illustrated by the fact that we can
instantiate the algebraic graph locale with the unit type.

This is an interesting direction for future work for a related reason that the
type class is interesting in Haskell: The type class forms the basis for 
a polymorphic graph transformation library, whereas the locale can be
seen as a \textit{polymorphic} graph formalization library. Any lemma we can prove
using the locale axioms becomes available to any graph representation
instantiating it. This obviously becomes more interesting the more material is
available in the locale. Haslbeck et al.\ chose a similar approach to implement
and verify Kruskal's algorithm independent of the concrete graph
representation~\cite{Kruskal-AFP}.

\section{Deep Embedding}\label{sec:deep}
Up until this point we were mainly concerned with exploiting the algebraic
structure of graphs for polymorphic graph constructions and transformations, and
for formal verification. However, that the deep embedding of the
construction primitives in a datatype is interesting in itself. Recall the
definition from the introduction:
\begin{minted}{haskell}
  data Graph a = Empty
               | Vertex a
               | Overlay (Graph a) (Graph a)
               | Connect (Graph a) (Graph a)
\end{minted}
An instance for the type class can be defined easily, however, we cannot use the
derived \hs{Eq} instance for the datatype, as it would violate the graph axioms.
One solution is to fold the deep embedding into a concrete graph representation to
obtain a law-abiding \hs{Eq} instance:
\begin{minted}{haskell}
  fold :: Graph g => Graph (Vertex g) -> g
  fold Empty         = empty
  fold (Vertex  x  ) = vertex x
  fold (Overlay x y) = overlay (fold x) (fold y)
  fold (Connect x y) = connect (fold x) (fold y)

  instance Ord a => Eq (Graph a) where
    x == y = fold x == (fold y :: Relation a)
\end{minted}
This approach is also currently implemented in the library available on
\texttt{hackage}. However, this is not ideal for multiple reasons. Coupling the
datatype to some other graph representation merely for equality checks is
inconvenient at best. More interestingly algebraic graphs can compactly
represent dense graphs. Consider the graph \hs{clique [1..n]}, which is linear
in size in the deep embedding, while it is quadratic when stored as a
\hs{Relation}. Requiring to fully expand graphs to check equality voids this
advantage. A promising approach to alleviate this issue is the \textit{modular
  graph decomposition}~\cite{mcconnell2005linear}, as it allows to minimize
algebraic graph expressions. It remains an open question if this decomposition
can be computed efficiently directly on the deep embedding.

More interesting questions for future work come up in connection with the deep
embedding. Currently only the transformations presented with the type class, i.e.\
transpose, as well as anything that can be done using the corresponding versions
of \hs{gmap} and \hs{bind}, are implemented directly on the deep embedding. No
known efficient implementations of fundamental graph algorithms, like
depth-first-search, that work directly on the deep embedding are
known~\cite{mokhov2017algebraic}. The library therefore resorts to converting
algebraic graph expressions to an adjacency map and uses algorithms
for those.

An immediate application which can be more efficient in space using
algebraic graphs that comes to mind is search, however, only in best case
examples, i.e.\ where only a small part of the graph has to be expanded. This is
due to the fact that the neighborhood of a vertex in an algebraic graph
expression can be computed in linear (in the size of the graph expression) time
and space. We already saw that the algebraic graph representation can be
significantly smaller than conventional representations.

Some benchmarks were performed (on an earlier version of the library) to compare
its performance to \texttt{containers} and \texttt{fgl} (and one other
representation). The results are mixed; depending on the area of application,
algebraic graphs may be competitive in performance, or in some cases, orders of
magnitudes slower. For the full results refer to the benchmark suite
repository\footnote{available at
  \texttt{\href{https://github.com/haskell-perf/graphs}{https://github.com/haskell-perf/graphs}}}.
\subsection{Verification}\label{sec:quotient}
We were also interested in using the deep embedding in formal verification using
Isabelle/HOL. Hence, we formalized most of the material from the library. This
feat turned out nicely, as the inductive datatype definition of the deep
embedding enables proofs by induction.

In the Haskell implementation of the deep embedding we already saw that
structural equality does not satisfy the axioms. There we can define a custom
\hs{Eq} instance to obtain a law-abiding datatype. In Isabelle/HOL we can follow
a similar approach, which results in a \texttt{quotient\_type}, which can be
thought of as an equivalence class construction on the raw datatype. For the
moment the equivalence is based on the tuple representation by Noschinski
mentioned in~\autoref{sec:verification}. Again, the minimization of algebraic
graph expressions for an equivalence relation on algebraic graphs, decoupled
from any other graph representation, is desirable.

The quotient type also allows us to formally prove that it satisfies the axioms,
by instantiating the locale. It also paves the way for further formalization
efforts on the deep embedding. However, this only becomes really interesting
with more future work, i.e.\ if algorithms that work directly on the deep
embedding or other advanced usages of it can be found (in other words, if there
is more material to formalize).

\section{Conclusion}\label{sec:conclusion}
We saw a novel approach to representing graphs, which is especially suitable for
functional programming. It allows for safe, polymorphic and elegant graph
constructions and transformations. Algebraic graphs also lend themselves nicely
to formal verification.

The alert reader might have noticed that the given primitives can only be used
to represent unlabeled graphs. Mokhov extended the algebraic graph library to
also represent edge-labeled graphs. There is no published material on this yet
(besides the source code), for more information refer to the talk by Mokhov
during Haskell eXchange 2018\footnote{available at
  \texttt{\href{https://skillsmatter.com/skillscasts/12361-labelled-algebraic-graphs}{https://skillsmatter.com/skillscasts/12361-labelled-algebraic-graphs}}}.

Algebraic graphs as presented have been successfully used in practice, e.g.\ for
asynchronous circuit design~\cite{beaumont2017high} and acceleration of graph
processing using FPGAs~\cite{mokhov2019language}. These are examples for which
the available transformations are sufficient. An important direction for future
research is going beyond those (relatively simple) transformations. That means
to find and formulate algorithms either inside the core type class or directly
on the deep embedding. Berghammer et. al~\cite{berghammer2020relational}
successfully employ algebraic approaches to formally verify graph algorithms.
However, they are using an algebra of relations as opposed to the one here
defined using graph construction primitives. Dolan~\cite{dolan2013fun} uses an
algebra on adjacency matrices to find paths in graphs.
Mokhov himself postulates that the algebraic approach allows to formulate graph
algorithms as systems of equations with unknowns. This approach may also be a
way to find novel graph algorithms.

\printbibliography

\end{document}
